# Goodness of fit of networks models to microbiome

## Required packages

### NetCoMi

```{r}
install.packages("devtools")
install.packages("BiocManager")
devtools::install_github("zdk123/pulsar")

# Since two of NetCoMi's dependencies are only available on GitHub, 
# it is recommended to install them first:
devtools::install_github("zdk123/SpiecEasi")
options(timeout = 600)
devtools::install_github("GraceYoon/SPRING")

# Install NetCoMi
devtools::install_github("stefpeschel/NetCoMi", 
                         repos = c("https://cloud.r-project.org/",
                                   BiocManager::repositories()))
```

### ggNetView

```{r}
# install.packages("devtools")
devtools::install_github("Jiawang1209/ggNetView")
```

## Reading network and counting nodes and edges 

### Network 1: Steinway et al. 2015

We read a network from a research paper where the data is given as an adjacency matrix, where rows and columns are microbial taxa. Non-zero values indicate interactions between taxa.

Link to data: https://bitbucket.org/gutmicrobiomepaper/microbiomenetworkmodelpaper/src/master/Metabolic_and_Experimental_Data/

Downloaded paths_in_boolean_net.txt and changed name to network_matrix.csv

Convert the matrix into a binary network (edge exists or not).
We then create a directed graph to obtain basic properties like the number of nodes and edges.

```{r}
library(igraph)

# Read the network matrix
mat <- as.matrix(read.csv("data/01-steinway/network_matrix.csv", row.names = 1))

# Look at the matrix to check it loaded correctly
mat

# Make a copy of the matrix
mat_binary <- mat

# Convert all non-zero values to 1
# Keep only where an edge exists
mat_binary[mat_binary != 0] <- 1

# Create the network
g_steinway <- graph_from_adjacency_matrix(mat_binary, mode = "directed")

# Count nodes
vcount(g_steinway)

# Count edges
ecount(g_steinway)

# Plot network
plot(g_steinway)
```

### Network 2: Hall et al. 2019

Link to data: https://www.ncbi.nlm.nih.gov/sra/PRJNA577760

```{r}

```

### Network 3: Liu et al. 2023

Link to data: https://github.com/YimengLiu9425/code/blob/master/study_code_Python/OTU%20table.xlsx

```{r}

```

### Network 4: Beghini et al. 2024

Link to data: https://github.com/human-nature-lab/strain_sharing/tree/main/data

```{r}

```

### Network 5: Hernandez et al. 2025

We read a microbial association network from a research paper where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/clayton-lab/marm_microbiome_metabolome_manuscript/blob/main/results/node_properties/associations.tsv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges <- read.delim("data/05-hernandez/associations.tsv")

# Look at the first few associations
head(edges)

# Create the network from the edge list
# X_features and Y_features are the connected taxa
# association is stored as an edge weight
g_hernandez <- graph_from_data_frame(
  d = edges[, c("X_features",
                "Y_features",
                "association")],
  directed = FALSE
)

# Count nodes
vcount(g_hernandez)

# Count edges
ecount(g_hernandez)

# Plot network
plot(g_hernandez)
```

### Network 6: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_mhno.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos1 <- read.csv("data/06-lacruz-pleguezuelos/edgeweights_mhno.csv" )

# Look at the first few associations
head(edges_lacruz_pleguezuelos1)

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos1 <- graph_from_data_frame( d = edges_lacruz_pleguezuelos1[, c("v1", "v2", "edgeweight")], directed = FALSE )

# Count nodes
vcount(g_lacruz_pleguezuelos1)

# Count edges
ecount(g_lacruz_pleguezuelos1)

# Plot network
plot(g_lacruz_pleguezuelos1)
```

### Network 7: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_muno.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos2 <- read.csv("data/07-lacruz-pleguezuelos/edgeweights_muno.csv" )

# Look at the first few associations
head(edges_lacruz_pleguezuelos2)

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos2 <- graph_from_data_frame( d = edges_lacruz_pleguezuelos2[, c("v1", "v2", "edgeweight")], directed = FALSE )

# Count nodes
vcount(g_lacruz_pleguezuelos2)

# Count edges
ecount(g_lacruz_pleguezuelos2)

# Plot network
plot(g_lacruz_pleguezuelos2)
```

### Network 8: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_muo.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos3 <- read.csv("data/08-lacruz-pleguezuelos/edgeweights_muo.csv" )

# Look at the first few associations
head(edges_lacruz_pleguezuelos3)

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos3 <- graph_from_data_frame( d = edges_lacruz_pleguezuelos3[, c("v1", "v2", "edgeweight")], directed = FALSE )

# Count nodes
vcount(g_lacruz_pleguezuelos3)

# Count edges
ecount(g_lacruz_pleguezuelos3)

# Plot network
plot(g_lacruz_pleguezuelos3)
```

### Network 9: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_v1.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos4 <- read.csv("data/09-lacruz-pleguezuelos/edgeweights_v1.csv" )

# Look at the first few associations
head(edges_lacruz_pleguezuelos4)

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos4 <- graph_from_data_frame( d = edges_lacruz_pleguezuelos4[, c("v1", "v2", "edgeweight")], directed = FALSE )

# Count nodes
vcount(g_lacruz_pleguezuelos4)

# Count edges
ecount(g_lacruz_pleguezuelos4)

# Plot network
plot(g_lacruz_pleguezuelos4)
```

### Network 10: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_v3.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos5 <- read.csv("data/10-lacruz-pleguezuelos/edgeweights_v3.csv" )

# Look at the first few associations
head(edges_lacruz_pleguezuelos5)

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos5 <- graph_from_data_frame( d = edges_lacruz_pleguezuelos5[, c("v1", "v2", "edgeweight")], directed = FALSE )

# Count nodes
vcount(g_lacruz_pleguezuelos5)

# Count edges
ecount(g_lacruz_pleguezuelos5)

# Plot network
plot(g_lacruz_pleguezuelos5)
```

### Network 15: Lugo-Martinez et al. 2019

We read an infant gut microbiome network from a research paper where the data is given in GraphML format. The network was generated from longitudinal microbiome data using a dynamic Bayesian network approach.

Link to data: https://github.com/jlugomar/longitudinal_microbiome_analysis_public/blob/master/visualization/infant_gut_microbiota_dbn_sample_alignment_filtered_sr3d_dbnIntraBoot_noPeriodOfStudy.graphml

The GraphML file is read directly into R as an `igraph` object. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read GraphML network
net_lugo_martinez <- read_graph("data/15-lugo-martinez/infant_gut_microbiota_dbn_sample_alignment_filtered_sr3d_dbnIntraBoot_noPeriodOfStudy.graphml",
                 format = "graphml")
vcount(net_lugo_martinez)
ecount(net_lugo_martinez)
# Count nodes
vcount(net_lugo_martinez)

# Count edges
ecount(net_lugo_martinez)

# Plot network
plot(net_lugo_martinez)
```

### Network 18: Ruiz-Perez et al. 2021

We read a network from a research paper where the data is given in GraphML format. The network contains microbial taxa, genes, metabolites, and host features as nodes. Edges represent inferred relationships from a Dynamic Bayesian Network.

Link to data: https://github.com/DaniRuizPerez/PALM-Public-Respository/blob/master/Networks/HEGTM/3Parents/human_ibd_microbiota_genes_metabolites_dbn_sample_alignment_sr14d_top100x100_host_taxa_reference_dbnIntraBoot.graphml

For this first analysis, we keep only the microbial taxa nodes so that this network is more comparable to the other microbiome networks in the project.

```{r}
library(igraph)

# Read GraphML network
net_ruiz_perez <- read_graph("data/18-ruiz-perez/human_ibd_microbiota_genes_metabolites_dbn_sample_alignment_sr14d_top100x100_host_taxa_reference_dbnIntraBoot.graphml",
                 format = "graphml")
vcount(net_ruiz_perez)
ecount(net_ruiz_perez)
# Count nodes
vcount(net_ruiz_perez)

# Count edges
ecount(net_ruiz_perez)

# Plot network
plot(net_ruiz_perez)

# TODO: Keep only microbial taxa nodes
# TODO: Create microbiome-only network
```

## Compute Degree Distribution

Degree measures how many connections each taxon has. This is one of the main features we will compare between microbiome networks and the simulated network models.

```{r}
# Compute degree
deg <- degree(g)
deg

summary(deg)

# Plot degree distribution
hist(deg, main = "Degree distribution", xlab = "Degree")
```

## Compute Clustering Coefficient

The clustering coefficient measures how connected neighboring taxa are to each other. This helps identify whether the microbiome network contains groups or clusters of highly connected taxa.

```{r}
# Compute clustering coefficient
clustering <- transitivity(g, type = "local", isolates = "zero")

clustering

summary(clustering)

# Plot clustering coefficient distribution
hist(clustering,
     main = "Clustering coefficient distribution",
     xlab = "Clustering coefficient",
     ylab = "Number of nodes")
```

## Compare with a model network

We will generate a random network (Erdos-Renyi) and compare it to the microbiome network above.

```{r}
# Generate an Erdos-Renyi network with the same number of nodes and edges
# Erdos-Renyi model network
g_er <- sample_gnm(vcount(g), ecount(g), directed = TRUE)

# Compute degree for the model network
deg_er <- degree(g_er)

# Compare distributions
hist(deg, col = "cyan",
     main = "Degree Comparison",
     xlab = "Degree",
     ylab = "Number of nodes")

hist(deg_er, col = "purple", add = TRUE)
```

```{r}
# Compute clustering coefficient for the model network
clustering_er <- transitivity(g_er,
                              type = "local",
                              isolates = "zero")

clustering_er

summary(clustering_er)

# Compare clustering coefficient distributions
hist(clustering,
     col = "cyan",
     main = "Clustering Coefficient Comparison",
     xlab = "Clustering coefficient",
     ylab = "Number of nodes")

hist(clustering_er,
     col = "purple",
     add = TRUE)
```
