# Goodness of fit of networks models to microbiome

## Required packages

### NetCoMi

```{r}

install.packages("devtools")
install.packages("BiocManager")
devtools::install_github("zdk123/pulsar")

# Since two of NetCoMi's dependencies are only available on GitHub, 
# it is recommended to install them first:
devtools::install_github("zdk123/SpiecEasi")
options(timeout = 600)
devtools::install_github("GraceYoon/SPRING")

# Install NetCoMi
devtools::install_github("stefpeschel/NetCoMi", 
                         repos = c("https://cloud.r-project.org/",
                                   BiocManager::repositories()))
```

### ggNetView

```{r}

# install.packages("devtools")
devtools::install_github("Jiawang1209/ggNetView")
```

## Reading network and counting nodes and edges 

### Network 1: Steinway et al 2015

We read a network from a research paper where the data is given as an adjacency matrix, where rows and columns are microbial taxa. Non-zero values indicate interactions between taxa.

Link to data: https://bitbucket.org/gutmicrobiomepaper/microbiomenetworkmodelpaper/src/master/Metabolic_and_Experimental_Data/

Downloaded paths_in_boolean_net.txt and changed name to network_matrix.csv

Convert the matrix into a binary network (edge exists or not).
We then create a directed graph to obtain basic properties like the number of nodes and edges.

```{r}
# Load package
library(igraph)

# Read the network matrix
mat <- as.matrix(read.csv("data/01-steinway/network_matrix.csv", row.names = 1))

# Look at the matrix to check it loaded correctly
mat

# Make a copy of the matrix
mat_binary <- mat

# Convert all non-zero values to 1
# Keep only where an edge exists
mat_binary[mat_binary != 0] <- 1

# Create the network
g <- graph_from_adjacency_matrix(mat_binary, mode = "directed")

# Count nodes
vcount(g)

# Count edges
ecount(g)

plot(g)
```

## Compute Degree Distribution

Degree measures how many connections each taxon has. This is one of the main features we will compare between microbiome networks and the simulated network models.

```{r}
# Compute degree
deg <- degree(g)
deg

summary(deg)

# Plot degree distribution
hist(deg, main = "Degree distribution", xlab = "Degree")
```

## Compute Clustering Coefficient

The clustering coefficient measures how connected neighboring taxa are to each other. This helps identify whether the microbiome network contains groups or clusters of highly connected taxa.

```{r}
# Compute clustering coefficient
clustering <- transitivity(g, type = "local", isolates = "zero")

clustering

summary(clustering)

# Plot clustering coefficient distribution
hist(clustering,
     main = "Clustering coefficient distribution",
     xlab = "Clustering coefficient",
     ylab = "Number of nodes")
```

## Compare with a model network

We will generate a random network (Erdos-Renyi) and compare it to the microbiome network above.

```{r}
# Generate an Erdos-Renyi network with the same number of nodes and edges
# Erdos-Renyi model network
g_er <- sample_gnm(vcount(g), ecount(g), directed = TRUE)

# Compute degree for the model network
deg_er <- degree(g_er)

# Compare distributions
hist(deg, col = "cyan",
     main = "Degree Comparison",
     xlab = "Degree",
     ylab = "Number of nodes")

hist(deg_er, col = "purple", add = TRUE)
```

```{r}
# Compute clustering coefficient for the model network
clustering_er <- transitivity(g_er,
                              type = "local",
                              isolates = "zero")

clustering_er

summary(clustering_er)

# Compare clustering coefficient distributions
hist(clustering,
     col = "cyan",
     main = "Clustering Coefficient Comparison",
     xlab = "Clustering coefficient",
     ylab = "Number of nodes")

hist(clustering_er,
     col = "purple",
     add = TRUE)
```
