# Goodness of fit of networks models to microbiome

## Required packages

### NetCoMi

```{r}
install.packages("devtools")
install.packages("BiocManager")
install.packages("rgexf")
devtools::install_github("zdk123/pulsar")

# Since two of NetCoMi's dependencies are only available on GitHub, 
# it is recommended to install them first:
devtools::install_github("zdk123/SpiecEasi")
options(timeout = 600)
devtools::install_github("GraceYoon/SPRING")

# Install NetCoMi
devtools::install_github("stefpeschel/NetCoMi", 
                         repos = c("https://cloud.r-project.org/",
                                   BiocManager::repositories()))
```

### ggNetView

```{r}
# install.packages("devtools")
devtools::install_github("Jiawang1209/ggNetView")
```

## Reading network and counting nodes and edges 

### Network 1: Steinway et al. 2015

We read a network from a research paper where the data is given as an adjacency matrix, where rows and columns are microbial taxa. Non-zero values indicate interactions between taxa.

Link to data: https://bitbucket.org/gutmicrobiomepaper/microbiomenetworkmodelpaper/src/master/Metabolic_and_Experimental_Data/

Downloaded paths_in_boolean_net.txt and changed name to network_matrix.csv

Convert the matrix into a binary network (edge exists or not).
We then create a directed graph to obtain basic properties like the number of nodes and edges.

```{r}
library(igraph)

# Read the network matrix
mat <- as.matrix(read.csv("data/01-steinway/network_matrix.csv", row.names = 1))

# Make a copy of the matrix
mat_binary <- mat

# Convert all non-zero values to 1
# Keep only where an edge exists
mat_binary[mat_binary != 0] <- 1

# Create the network
g_steinway <- graph_from_adjacency_matrix(mat_binary, mode = "directed")

# Count nodes
vcount(g_steinway)

# Count edges
ecount(g_steinway)

# Plot network
plot(g_steinway)
```

### Network 2: Hall et al. 2019

Link to data: https://www.ncbi.nlm.nih.gov/sra/PRJNA577760

```{r}

```

### Network 3: Liu et al. 2023

Link to data: https://github.com/YimengLiu9425/code/blob/master/study_code_Python/OTU%20table.xlsx

```{r}

```

### Network 4: Beghini et al. 2024

Link to data: https://github.com/human-nature-lab/strain_sharing/tree/main/data

```{r}

```

### Network 5: Hernandez et al. 2025

We read a microbial association network from a research paper where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/clayton-lab/marm_microbiome_metabolome_manuscript/blob/main/results/node_properties/associations.tsv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges <- read.delim("data/05-hernandez/associations.tsv")

# Look at the first few associations
head(edges)

# Create the network from the edge list
# X_features and Y_features are the connected taxa
# association is stored as an edge weight
g_hernandez <- graph_from_data_frame(d = edges[, c("X_features", "Y_features", "association")], directed = FALSE)

# Count nodes
vcount(g_hernandez)

# Count edges
ecount(g_hernandez)

# Plot network
plot(g_hernandez)
```

### Network 6: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_mhno.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos1 <- read.csv("data/06-lacruz-pleguezuelos/edgeweights_mhno.csv")

# Look at the first few associations
head(edges_lacruz_pleguezuelos1)

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos1 <- graph_from_data_frame(d = edges_lacruz_pleguezuelos1[, c("v1", "v2", "edgeweight")], directed = FALSE)

# Count nodes
vcount(g_lacruz_pleguezuelos1)

# Count edges
ecount(g_lacruz_pleguezuelos1)

# Plot network
plot(g_lacruz_pleguezuelos1)
```

### Network 7: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_muno.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos2 <- read.csv("data/07-lacruz-pleguezuelos/edgeweights_muno.csv")

# Look at the first few associations
head(edges_lacruz_pleguezuelos2)

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos2 <- graph_from_data_frame(d = edges_lacruz_pleguezuelos2[, c("v1", "v2", "edgeweight")], directed = FALSE)

# Count nodes
vcount(g_lacruz_pleguezuelos2)

# Count edges
ecount(g_lacruz_pleguezuelos2)

# Plot network
plot(g_lacruz_pleguezuelos2)
```

### Network 8: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_muo.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos3 <- read.csv("data/08-lacruz-pleguezuelos/edgeweights_muo.csv")

# Look at the first few associations
head(edges_lacruz_pleguezuelos3)

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos3 <- graph_from_data_frame(d = edges_lacruz_pleguezuelos3[, c("v1", "v2", "edgeweight")], directed = FALSE)

# Count nodes
vcount(g_lacruz_pleguezuelos3)

# Count edges
ecount(g_lacruz_pleguezuelos3)

# Plot network
plot(g_lacruz_pleguezuelos3)
```

### Network 9: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_v1.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos4 <- read.csv("data/09-lacruz-pleguezuelos/edgeweights_v1.csv")

# Look at the first few associations
head(edges_lacruz_pleguezuelos4)

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos4 <- graph_from_data_frame(d = edges_lacruz_pleguezuelos4[, c("v1", "v2", "edgeweight")], directed = FALSE)

# Count nodes
vcount(g_lacruz_pleguezuelos4)

# Count edges
ecount(g_lacruz_pleguezuelos4)

# Plot network
plot(g_lacruz_pleguezuelos4)
```

### Network 10: Lacruz-Pleguezuelos et al. 2026

We read a microbial association network from the MHOmicrobiome GitHub repository, where the data is provided as an edge list. Each row represents an association between two microbial taxa, along with the strength of that association.

Link to data: https://github.com/blacruz17/MHOmicrobiome/blob/main/data/edgeweights_v3.csv

The edge list is converted directly into an undirected graph. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read the edge list
edges_lacruz_pleguezuelos5 <- read.csv("data/10-lacruz-pleguezuelos/edgeweights_v3.csv")

# Create the network from the edge list
# v1 and v2 are the connected taxa
# edgeweight is stored as an edge weight
g_lacruz_pleguezuelos5 <- graph_from_data_frame(d = edges_lacruz_pleguezuelos5[, c("v1", "v2", "edgeweight")], directed = FALSE)

# Count nodes
vcount(g_lacruz_pleguezuelos5)

# Count edges
ecount(g_lacruz_pleguezuelos5)

# Plot network
plot(g_lacruz_pleguezuelos5)
```

### Network 11: Jeong et al. 2023

The original data was downloaded as a journal table CSV file. Because the journal table used merged-cells, the raw CSV files were cleaned into standard edge lists before being read into R. The cleaned edge list is then converted into an undirected graph. We then obtain the number of nodes and edges.

Link to data: https://www.tandfonline.com/doi/figure/10.1080/20002297.2023.2186591?scroll=top&needAccess=true

This table represents the compositional network between bacterial genera within the S type oral microbiome.

```{r}
library(igraph)

# Read the cleaned edge list
edges_jeong_s_type <- read.csv("data/11-jeong/jeong_s_type_edges.csv")

# Look at the first few associations
head(edges_jeong_s_type)

# Create the network from the edge list
# Source and target are the connected bacterial genera
# Weight is the SparCC R-value
# Sign indicates whether the association is positive or negative
g_jeong_s_type <- graph_from_data_frame(d = edges_jeong_s_type[, c("source", "target", "weight", "sign")], directed = FALSE)

# Count nodes
vcount(g_jeong_s_type)

# Count edges
ecount(g_jeong_s_type)

# Plot network
plot(g_jeong_s_type)
```

### Network 12: Jeong et al. 2023

The original data was downloaded as a journal table CSV file. Because the journal table used merged-cells, the raw CSV files were cleaned into standard edge lists before being read into R. The cleaned edge list is converted into an undirected graph. We then obtain the number of nodes and edges.

Link to data: https://www.tandfonline.com/doi/figure/10.1080/20002297.2023.2186591?scroll=top&needAccess=true

This table represents the compositional network between bacterial genera within the S type oral microbiome.

```{r}
library(igraph)

# Read the cleaned edge list
edges_jeong_h_type <- read.csv("data/12-jeong/jeong_h_type_edges.csv")

# Look at the first few associations
head(edges_jeong_h_type)

# Create the network from the edge list
# Source and target are the connected bacterial genera
# Weight is the SparCC R-value
# Sign indicates whether the association is positive or negative
g_jeong_h_type <- graph_from_data_frame(d = edges_jeong_h_type[,c("source","target", "weight", "sign")], directed = FALSE)

# Count nodes
vcount(g_jeong_h_type)

# Count edges
ecount(g_jeong_h_type)

# Plot network
plot(g_jeong_h_type)
```

### Network 13: Goyal et al. 2021

We read a human gut microbiome cross-feeding network from a research paper. The network connects microbial species and metabolites, where interactions represent metabolite consumption, production, or both. The original network was based on experimentally verified and manually curated gut microbiome interactions.

Link to data: https://github.com/maslov-group/ML_human_gut/blob/master/data/pruned_chia_network.csv

The CSV file is read into R as an edge list and converted into an undirected igraph object. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read cross-feeding edge list
net_goyal_data <- read.csv("data/13-goyal/pruned_chia_network.csv")

# Rename columns
names(net_goyal_data) <- c("metabolite_id", "microbe_id", "edge_type")

# Add prefixes so metabolite and microbe nodes are clearly distinct
net_goyal_data$metabolite_id <- paste0("metabolite_", net_goyal_data$metabolite_id)

net_goyal_data$microbe_id <- paste0("microbe_",nnet_goyal_data$microbe_id)

# Convert edge list into an undirected igraph object
net_goyal <- graph_from_data_frame(net_goyal_data, directed = FALSE)

# Count nodes
vcount(net_goyal)

# Count edges
ecount(net_goyal)

# Plot network
plot(net_goyal)
```

### Network 15: Lugo-Martinez et al. 2019

We read an infant gut microbiome network from a research paper where the data is given in GraphML format. The network was generated from longitudinal microbiome data using a dynamic Bayesian network approach.

Link to data: https://github.com/jlugomar/longitudinal_microbiome_analysis_public/blob/master/visualization/infant_gut_microbiota_dbn_sample_alignment_filtered_sr3d_dbnIntraBoot_noPeriodOfStudy.graphml

The GraphML file is read directly into R as an `igraph` object. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read GraphML network
net_lugo_martinez <- read_graph("data/15-lugo-martinez/infant_gut_microbiota_dbn_sample_alignment_filtered_sr3d_dbnIntraBoot_noPeriodOfStudy.graphml",
                                format = "graphml")

# Count nodes
vcount(net_lugo_martinez)

# Count edges
ecount(net_lugo_martinez)

# Plot network
plot(net_lugo_martinez)
```

### Network 17: Melnyk et al. 2020

```{r}

```


### Network 18: Ruiz-Perez et al. 2021

We read a network from a research paper where the data is given in GraphML format. The network contains microbial taxa, genes, metabolites, and host features as nodes. Edges represent inferred relationships from a Dynamic Bayesian Network.

Link to data: https://github.com/DaniRuizPerez/PALM-Public-Respository/blob/master/Networks/HEGTM/3Parents/human_ibd_microbiota_genes_metabolites_dbn_sample_alignment_sr14d_top100x100_host_taxa_reference_dbnIntraBoot.graphml

For this first analysis, we keep only the microbial taxa nodes so that this network is more comparable to the other microbiome networks in the project.

```{r}
library(igraph)

# Read GraphML network
net_ruiz_perez <- read_graph("data/18-ruiz-perez/human_ibd_microbiota_genes_metabolites_dbn_sample_alignment_sr14d_top100x100_host_taxa_reference_dbnIntraBoot.graphml",
                             format = "graphml")

# Keep only microbial taxa nodes
taxa_nodes <- V(net_ruiz_perez)[grepl("^s__", V(net_ruiz_perez)$id)]

# Create microbiome-only network
net_ruiz_perez_taxa <- induced_subgraph(net_ruiz_perez, vids = taxa_nodes)

# Count nodes
vcount(net_ruiz_perez_taxa)

# Count edges
ecount(net_ruiz_perez_taxa)

# Plot network
plot(net_ruiz_perez_taxa)
```

### Network 19: Lymberopoulos et al. 2022

We read a gut microbiome co-enrichment network from a research paper. The network was created from gut microbiome samples using Topological Data Analysis, where connections represent co-enrichment relationships between microbial taxa.

Link to data: https://github.com/thesharmalab-team/microbiome_covid/blob/main/CovidTDA/data_and_scripts/binarised_coenrichment_GMrepo_TDA_PD.csv

The CSV file is read as an adjacency matrix and converted into an undirected igraph network. We then obtain basic network properties such as the number of nodes and edges.

```{r}
library(igraph)

# Read adjacency matrix
gut_data <- read.csv("data/19-lymberopoulos/binarised_coenrichment_GMrepo_TDA_PD.csv", row.names = 1, check.names = FALSE)

# Keep only gut microbiome taxa
gut_data <- as.matrix(gut_data[1:985, 1:985])

# Create undirected network
net_lymberopoulos <- graph_from_adjacency_matrix(gut_data, mode = "undirected", diag = FALSE)

# Count nodes
vcount(net_lymberopoulos)

# Count edges
ecount(net_lymberopoulos)

# Plot network
plot(net_lymberopoulos)
```

### Network 20: Loftus et al. 2021

We read a colorectal cancer gut microbiome bacterial association network from a research paper. The data contains pairs of associated bacterial species and their association weights.

Link to data: https://github.com/Markloftus/CancerMicrobiome/blob/main/Files/Dataframes/OnlyCRC_associations.csv

The bacterial pairs are separated into two columns to create an edge list. The edge list is then converted into an undirected `igraph` network.

```{r}
library(igraph)

# Read data
crc_data <- read.csv("data/20-loftus/OnlyCRC_associations.csv")

# Create edge list
crc_edges <- data.frame(bacteria1 = sub("<-->.*", "", crc_data[[1]]),
                        bacteria2 = sub(".*<-->", "", crc_data[[1]]),
                        weight = crc_data[[2]])

# Create network
net_loftus_crc <- graph_from_data_frame(crc_edges, directed = FALSE)

# Count nodes
vcount(net_loftus_crc)

# Count edges
ecount(net_loftus_crc)

# Plot network
plot(net_loftus_crc)
```

### Network 21: Loftus et al. 2021

We read a healthy gut microbiome bacterial association network from a research paper. The data contains pairs of associated bacterial species and their association weights.

Link to data: https://github.com/Markloftus/CancerMicrobiome/blob/main/Files/Dataframes/OnlyHealthy_Associations.csv

The bacterial pairs are separated into two columns to create an edge list. The edge list is then converted into an undirected `igraph` network.

```{r}
library(igraph)

# Read data
healthy_data <- read.csv("data/21-loftus/OnlyHealthy_Associations.csv")

# Create edge list
healthy_edges <- data.frame(bacteria1 = sub("<-->.*", "", healthy_data[[1]]),
                            bacteria2 = sub(".*<-->", "", healthy_data[[1]]),
                            weight = healthy_data[[2]])

# Create network
net_loftus_healthy <- graph_from_data_frame(healthy_edges, directed = FALSE)

# Count nodes
vcount(net_loftus_healthy)

# Count edges
ecount(net_loftus_healthy)

# Plot network
plot(net_loftus_healthy)
```

### Network 22: Ravikrishnan et al. 2021

We read a gut microbiome metabolic association network under DMEM conditions. The data contains pairs of gut microbial species and their Metabolic Support Index (MSI) values.

Link to data: https://github.com/aarthi31/gutmicrobiome/blob/master/MSIResults/edges_msi_benefit_dmem_withedgeval.txt

The microbial pairs are used as an edge list, with the MSI value as the edge weight. The edge list is then converted into a directed `igraph` network.

```{r}
library(igraph)

# Read DMEM interaction data
dmem_data <- read.table("data/22-ravikrishnan/edges_msi_benefit_dmem_withedgeval.txt", header = TRUE, sep = "\t")

# Create directed edge list
dmem_edges <- data.frame(from = dmem_data[[2]], to = dmem_data[[1]],weight = dmem_data[[3]])

# Create directed network
net_ravikrishnan_dmem <- graph_from_data_frame(dmem_edges, directed = TRUE)

# Count nodes
vcount(net_ravikrishnan_dmem)

# Count edges
ecount(net_ravikrishnan_dmem)

# Plot network
plot(net_ravikrishnan_dmem)
```

## Network 23: Ravikrishnan et al. 2021

We read a gut microbiome metabolic association network under high-fibre conditions. The data contains pairs of gut microbial species and their Metabolic Support Index (MSI) values.

Link to data: https://github.com/aarthi31/gutmicrobiome/blob/master/MSIResults/edges_msi_benefit_hf_withedgeval.txt

The microbial pairs are used as an edge list, with the MSI value as the edge weight. The edge list is then converted into a directed `igraph` network.

```{r}
library(igraph)

# Read high-fibre interaction data
hf_data <- read.table("data/23-ravikrishnan/edges_msi_benefit_hf_withedgeval.txt", header = TRUE, sep = "\t")

# Create directed edge list
hf_edges <- data.frame(from = hf_data[[2]], to = hf_data[[1]], weight = hf_data[[3]])

# Create directed network
net_ravikrishnan_hf <- graph_from_data_frame(hf_edges, directed = TRUE)

# Count nodes
vcount(net_ravikrishnan_hf)

# Count edges
ecount(net_ravikrishnan_hf)

# Plot network
plot(net_ravikrishnan_hf)
```

## Network 27: Li et al. 2019

We read a healthy human gut microbiome interaction network from a research paper. The network contains microbial species as nodes and directed Transfer Entropy relationships between species as edges.

Link to data: https://github.com/HokudaiNexusLab/Microbiome/tree/master/resultsNetwork

The node and edge CSV files are read into R and combined to create the network. The edge weights represent Transfer Entropy, which measures information flow between microbial species. The files are then converted into a directed `igraph` network.

```{r}
library(igraph)

# Read nodes
healthy_nodes <- read.csv("data/Nodes_Healthy_Group_withoutC_C1.csv")

# Read edges
healthy_edges <- read.csv("data/Edges_Healthy_Group_withoutC_C1.csv")

# Create network
net_healthy <- graph_from_data_frame(healthy_edges, directed = TRUE, vertices = healthy_nodes)

# Count nodes
vcount(net_healthy)

# Count edges
ecount(net_healthy)

# Plot network
plot(net_healthy, vertex.label = V(net_healthy)$Label)
```

## Network 28: Li et al. 2019

We read an unhealthy IBS human gut microbiome interaction network from a research paper. The network contains microbial species as nodes and directed Transfer Entropy relationships between species as edges. The paper compares interaction networks across healthy and unhealthy microbiome states.

Link to data: https://github.com/HokudaiNexusLab/Microbiome/tree/master/resultsNetwork

The node and edge CSV files are read into R and combined to create the network. The edge weights represent Transfer Entropy, which measures information flow between microbial species. The files are then converted into a directed igraph network.

```{r}
# Read nodes
unhealthy_nodes <- read.csv("data/Nodes_MaxEnt_UnHealthy.csv")

# Read edges
unhealthy_edges <- read.csv("data/Edges_MaxEnt_UnHealthy.csv")

# Create network
net_unhealthy <- graph_from_data_frame(unhealthy_edges, directed = TRUE, vertices = unhealthy_nodes)

# Count nodes
vcount(net_unhealthy)

# Count edges
ecount(net_unhealthy)

# Plot network
plot(net_unhealthy, vertex.label = V(net_unhealthy)$Label)
```

## Compute Degree Distribution

Degree measures how many connections each taxon has. This is one of the main features we will compare between microbiome networks and the simulated network models.

```{r}
# Compute degree
deg <- degree(g)
deg

summary(deg)

# Plot degree distribution
hist(deg, main = "Degree distribution", xlab = "Degree")
```

## Compute Clustering Coefficient

The clustering coefficient measures how connected neighboring taxa are to each other. This helps identify whether the microbiome network contains groups or clusters of highly connected taxa.

```{r}
# Compute clustering coefficient
clustering <- transitivity(g, type = "local", isolates = "zero")

clustering

summary(clustering)

# Plot clustering coefficient distribution
hist(clustering,
     main = "Clustering coefficient distribution",
     xlab = "Clustering coefficient",
     ylab = "Number of nodes")
```

## Compare with a model network

We will generate a random network (Erdos-Renyi) and compare it to the microbiome network above.

```{r}
# Generate an Erdos-Renyi network with the same number of nodes and edges
# Erdos-Renyi model network
g_er <- sample_gnm(vcount(g), ecount(g), directed = TRUE)

# Compute degree for the model network
deg_er <- degree(g_er)

# Compare distributions
hist(deg, col = "cyan",
     main = "Degree Comparison",
     xlab = "Degree",
     ylab = "Number of nodes")

hist(deg_er, col = "purple", add = TRUE)
```

```{r}
# Compute clustering coefficient for the model network
clustering_er <- transitivity(g_er,
                              type = "local",
                              isolates = "zero")

clustering_er

summary(clustering_er)

# Compare clustering coefficient distributions
hist(clustering,
     col = "cyan",
     main = "Clustering Coefficient Comparison",
     xlab = "Clustering coefficient",
     ylab = "Number of nodes")

hist(clustering_er,
     col = "purple",
     add = TRUE)
```
